### 2022-06-06 src/ms.tex
making progress on the stats section

% Custom command
\newcommand{\ud}{\,\mathrm{d}}
\renewcommand{\vec}[1]{\bm{#1}}

@@

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\boldsymbol{\theta}_S$ to be the angular position of the source star
on the plane of the sky and $\boldsymbol{\mu}_L$ to be its proper motion
vector. Likewise for the lens. We thus have
\begin{align}
    \boldsymbol{\theta}_S(t) & =\boldsymbol{\theta}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}                              \\
    \boldsymbol{\theta}_L(t) & =\boldsymbol{\theta}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\boldsymbol{\theta}_L(t)
        -\boldsymbol{\theta}_S(t)}{\theta_E}=
    \frac{\boldsymbol{\theta}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\boldsymbol{\theta}_{L S, 0}\equiv\boldsymbol{\theta}_{L,0}-\boldsymbol{\theta}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\boldsymbol{\theta}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\boldsymbol{\theta}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The magnification is then a function of three parameters $(t_0, u_0, t_E)$
and the resulting curve as a function of time is often called the Paczy\'nski curve
\citep{1986ApJ...304....1P,1986ApJ...301..503P}.
The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:paczynski_curve}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$).
In the limits of $u_0\rightarrow 0$ and $u_0\gg 1$  there is a continuous
mathematical degeneracy between the parameters $t_0$, $u_0$ and $t_E$
\citep{1997ApJ...487...55W}.

@@

In this section I will describe the annual parallax effect. We start by
modifying Equations~\ref{eq:source_position} and \ref{eq:lens_position} to
include the projected motion of the Sun relative to the Earth on the plane of
the sky. We project $\mathbf s$ -- the position vector of the Sun relative to
Earth, onto a plane perpendicular to the line of sight towards the source star
(plane of the sky) which is defined by the unit vector $\boldsymbol{\hat n}$
normal to the plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky
coordinates of the source star $(\alpha,\delta)$ (right ascension and
declination). We choose to work in geocentric equatorial coordinates defined by
the spherical unit vectors $\hat{\mathbf e}_n$, pointing North, and
$\hat{\mathbf e}_e$ pointing eastward such that the coordinate system is
right-handed. The unit vectors are defined by
\begin{align}
    \hat{\mathbf e}_e & = \hat{\mathbf z}\times \hat{\mathbf n}  \\
    \hat{\mathbf e}_n & = \hat{\mathbf n}\times\hat{\mathbf e}_e
\end{align}
The two components of $\mathbf s$ projected onto the plane of the sky are then
\begin{align}
    \zeta_e(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_e \\
    \zeta_n(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_n
\end{align}
In practice, we can retrieve $\mathbf{s}$ using NASA's JPL Horizons system and compute
the projected separation at any time $t$.
The angular positions of the source and the lens
(Equations~\ref{eq:source_position} and \ref{eq:lens_position}) then become
\begin{align}
    \boldsymbol{\theta}_S(t) & = \boldsymbol{\theta}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
    +\pi_S\,\boldsymbol{\zeta}(t)                                                    \\
    \boldsymbol{\theta}_L(t) & = \boldsymbol{\theta}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
    +\pi_L\,\boldsymbol{\zeta}(t)
\end{align}
where  $\pi_S\equiv 1\,\mathrm{au}/D_S$ is the source parallax, and $\pi_L$ is the
lens parallax.
The relative separation is then
\begin{equation}
    \boldsymbol{u}(t)= \frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
    +\frac{t-t_0}{\theta_E}\boldsymbol{\mu}_{LS}+\pi_{E}\,\boldsymbol{\zeta}(t)
    \label{eq:relative_separation_parallax}
\end{equation}
where
\begin{equation}
    \pi_E\equiv \frac{\pi_{LS}}{\theta_E}
    \label{eq:pi_E}
\end{equation}

@@

\subsection{A system with $N$ lenses}
Microlensing with more than one lens is considerably more complex than the case
of a single lens. Consider a system with $N$ point mass lenses with a total
mass $M=\sum_{i=1}^Nm_i$. As before, we have the angular source position in the
source plane, given by the the dimensionless vector
$\mathbf{y}=\bm\beta/(D_\textrm{S}\theta_E)$ and the position of the images in
the lens plane, given by given by
$\mathbf{x}=\bm\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now refers to
the angular Einstein radius corresponding to the total mass $M$. The lens
equation then contains a sum over the deflection angles $\bm\alpha_i$
corresponding to each point mass $m_i$\footnote{This is because we are working
    within the linearized GR framework so the deflection angles are additive.}:
\begin{equation}
    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\bm\alpha_i(\mathbf{x},\mathbf{x}_i)
\end{equation}
where $\bm\alpha_i$ is the deflection angle due to the $i$th lens.
Using Equation~\ref{eq:lens_equation_dimensionless}, we obtain
\begin{equation}
    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{x}-\mathbf{x}_i}
    {\lvert \mathbf{x}-\mathbf{x}_i\rvert^2}
    \label{eq:lens_equation_general}
\end{equation}
where $\epsilon_i\equiv m_i/M$.

@@

\section{Occultation and phase curve mapping}
\label{sec:occultations}
In this section I will briefly review the theory behind modeling phase curves and
occultation light curves of spherical bodies for the case of emitted light
(isotropic blackbody emission from the body itself) and reflected light
(starlight scattered from the surface or atmosphere of the body). The following is
mostly a short summary of the papers introducing the \textsf{starry} framework
\citep{2019AJ....157...64L,2021arXiv210306275L}.
\subsection{The Starry framework}
\begin{figure}[t]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/spherical_harmonics.pdf}
        \caption{Real spherical harmonics up to degree $l=5$ computed from
            Equation~\ref{eq:spherical_harmonics}. Figure adapted from Figure 1 in
            \citet{2019AJ....157...64L}.}
        \label{fig:spherical_harmonics}
    \end{centering}
\end{figure}
\subsubsection{Emitted light}
Consider a spherical body of unit radius emitting light isotropically (a
Lambertian emitter) at each point $(\theta,\phi)$ where $\theta$ is the
inclination angle and $\phi$ is the azimuthal angle. To compute a phase curve
or an occultation light curve we need to be able to integrate the flux over the
entire projected disc of the body given the distribution of specific intensity
$I(\theta, \phi)$, the 3D orientation of the body in space, and the location of
a possible occultor relative to the body. These two-dimensional integrals can
always be computed numerically (see for instance the approaches presented in
\citet{2018AJ....156..146F,2018MNRAS.477.2613L}) but
\citet{2019AJ....157...64L} showed that it is possible to compute them
analytically if the specific intensity is first expanded in a basis of
spherical harmonics and then mapped to a different basis more suitable for
evaluating the integrals.

@@

\section{Bayesian statistics}
\label{sec:bayes}
Having covered the basic physics behind gravitational lensing and occultation mapping I
now turn to the subject of statistical inference in general and more focus more
specifically on Bayesian statistics in particular because it provides a straightforward
and elegant theoretical framework for solving inverse problems of the kind we mentioned
in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}.

\subsection{Probability theory}
\subsection{The meaning of probability}
\subsection{Probability as frequency of events in repeated trials}
\subsection{Probability as degrees of belief}
\subsection{The likelihood function}
% mention the likelihood principle
\subsection{Priors}
\subsection{On machine learning and the difference between explanation and prediction}
% mention that even if we have a model focused on inference, it might be best if we judge 
% its quality through its predictive performance

\section{Inference}
\label{sec:inference}
\subsection{The curse of dimensionality}
% check that paper on physics and ML by that MIT guy, there's a good stuff there. 
% check also Ian Murray's lecture notes  on MLPR

\subsection{There's no free lunch}
% exploration vs. explotation 
\subsection{Optimization vs. sampling}
\subsection{Sampling and optimization methods}
\subsubsection{Maximum likelihood}
\subsubsection{The Laplace approximation}
\subsubsection{Variational inference}
\subsubsection{Importance sampling}
\subsubsection{Rejection sampling}
\subsubsection{Markov Chain Monte Carlo}
\subsubsection{Hamiltonian Monte Carlo}
\subsubsection{Other notable methods}

\section{Model validation and comparison}
\label{sec:model_comparison}

``Furthermore, in the future, a blind search for planetary signals will not
turn up all potential competing models unless those models
are specifically considered.''\citep{2017AJ....153..129J}.
\citet{2021AJ....162...59R}
\citet{2020AJ....160...17H}

\subsection{Cross validation}
% discuss vanilla k-fold CV, stress out just how important and central CV is to science
\subsubsection{Pareto smoothed importance sampling LOO}
\subsection{Model comparison}
% don't do it!

\section{Putting it all together -- \emph{probabilistic programming}}
\label{sec:programming}
\subsection{The success of machine learning}
% that quote from Yann LeCun on differentiable code 
\subsection{Automatic differentiation}
\subsubsection{Forward mode autodiff}
\subsubsection{Reverse mode autodiff or backpropagation}
\subsubsection{Jacobian vector products}
\subsection{JAX}
\subsection{Numpyro}

>>>>

% Custom command
\newcommand{\ud}{\,\mathrm{d}}
\renewcommand{\vec}[1]{\boldsymbol{\mathbf{#1}}}

@@

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\vec{x}_S$ to be the angular position of the source star on the plane
of the sky and $\boldsymbol{\mu}_L$ to be its proper motion vector. Likewise
for the lens. We thus have
\begin{align}
    \vec{x}_S(t) & =\vec{x}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}      \\
    \vec{x}_L(t) & =\vec{x}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\vec{x}_L(t)
        -\vec{x}_S(t)}{\theta_E}=
    \frac{\vec{x}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\vec{x}_{L S, 0}\equiv\vec{x}_{L,0}-\vec{x}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\vec{x}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\vec{x}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The magnification is then a function of three parameters $(t_0, u_0, t_E)$
and the resulting curve as a function of time is often called the Paczy\'nski curve
\citep{1986ApJ...304....1P,1986ApJ...301..503P}.
The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:paczynski_curve}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$).
In the limits of $u_0\rightarrow 0$ and $u_0\gg 1$  there is a continuous
mathematical degeneracy between the parameters $t_0$, $u_0$ and $t_E$
\citep{1997ApJ...487...55W}.

@@

In this section I will describe the annual parallax effect. We start by
modifying Equations~\ref{eq:source_position} and \ref{eq:lens_position} to
include the projected motion of the Sun relative to the Earth on the plane of
the sky. We project $\mathbf s$ -- the position vector of the Sun relative to
Earth, onto a plane perpendicular to the line of sight towards the source star
(plane of the sky) which is defined by the unit vector $\boldsymbol{\hat n}$
normal to the plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky
coordinates of the source star $(\alpha,\delta)$ (right ascension and
declination). We choose to work in geocentric equatorial coordinates defined by
the spherical unit vectors $\hat{\mathbf e}_n$, pointing North, and
$\hat{\mathbf e}_e$ pointing eastward such that the coordinate system is
right-handed. The unit vectors are defined by
\begin{align}
    \hat{\mathbf e}_e & = \hat{\mathbf z}\times \hat{\mathbf n}  \\
    \hat{\mathbf e}_n & = \hat{\mathbf n}\times\hat{\mathbf e}_e
\end{align}
The two components of $\mathbf s$ projected onto the plane of the sky are then
\begin{align}
    \zeta_e(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_e \\
    \zeta_n(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_n
\end{align}
In practice, we can retrieve $\mathbf{s}$ using NASA's JPL Horizons system and compute
the projected separation at any time $t$.
The angular positions of the source and the lens
(Equations~\ref{eq:source_position} and \ref{eq:lens_position}) then become
\begin{align}
    \vec{x}_S(t) & = \vec{x}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
    +\pi_S\,\boldsymbol{\zeta}(t)                            \\
    \vec{x}_L(t) & = \vec{x}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
    +\pi_L\,\boldsymbol{\zeta}(t)
\end{align}
where  $\pi_S\equiv 1\,\mathrm{au}/D_S$ is the source parallax, and $\pi_L$ is the
lens parallax.
The relative separation is then
\begin{equation}
    \boldsymbol{u}(t)= \frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
    +\frac{t-t_0}{\theta_E}\boldsymbol{\mu}_{LS}+\pi_{E}\,\boldsymbol{\zeta}(t)
    \label{eq:relative_separation_parallax}
\end{equation}
where
\begin{equation}
    \pi_E\equiv \frac{\pi_{LS}}{\theta_E}
    \label{eq:pi_E}
\end{equation}

@@

\subsection{A system with $N$ lenses}
Microlensing with more than one lens is considerably more complex than the case
of a single lens. Consider a system with $N$ point mass lenses with a total
mass $M=\sum_{i=1}^Nm_i$. As before, we have the angular source position in the
source plane, given by the the dimensionless vector
$\mathbf{y}=\boldsymbol\beta/(D_\textrm{S}\theta_E)$ and the position of the
images in the lens plane, given by given by
$\mathbf{x}=\boldsymbol\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now
refers to the angular Einstein radius corresponding to the total mass $M$. The
lens equation then contains a sum over the deflection angles
$\boldsymbol\alpha_i$ corresponding to each point mass $m_i$\footnote{This is
    because we are working within the linearized GR framework so the deflection
    angles are additive.}:
\begin{equation}
    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\boldsymbol\alpha_i(\mathbf{x},\mathbf{x}_i)
\end{equation}
where $\boldsymbol\alpha_i$ is the deflection angle due to the $i$th lens.
Using Equation~\ref{eq:lens_equation_dimensionless}, we obtain
\begin{equation}
    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{x}-\mathbf{x}_i}
    {\lvert \mathbf{x}-\mathbf{x}_i\rvert^2}
    \label{eq:lens_equation_general}
\end{equation}
where $\epsilon_i\equiv m_i/M$.

@@

\section{Occultation and phase curve mapping}
\label{sec:occultations}
In this section I will briefly review the theory behind modeling phase curves and
occultation light curves of spherical bodies for the case of emitted light
(isotropic blackbody emission from the body itself) and reflected light
(starlight scattered from the surface or atmosphere of the body). The following is
mostly a short summary of the papers introducing the \textsf{starry} framework
\citep{2019AJ....157...64L,2021arXiv210306275L}.
\subsection{The starry algorithm}
\begin{figure}[t]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/spherical_harmonics.pdf}
        \caption{Real spherical harmonics up to degree $l=5$ computed from
            Equation~\ref{eq:spherical_harmonics}. Figure adapted from Figure 1 in
            \citet{2019AJ....157...64L}.}
        \label{fig:spherical_harmonics}
    \end{centering}
\end{figure}
\subsubsection{Emitted light}
Consider a spherical body of unit radius emitting light isotropically (a
Lambertian emitter) at each point $(\theta,\phi)$ where $\theta$ is the
inclination angle and $\phi$ is the azimuthal angle. To compute a phase curve
or an occultation light curve we need to be able to integrate the flux over the
entire projected disc of the body given the distribution of specific intensity
$I(\theta, \phi)$, the 3D orientation of the body in space, and the location of
a possible occultor relative to the body. These two-dimensional integrals can
always be computed numerically (see for instance the approaches presented in
\citet{2018AJ....156..146F,2018MNRAS.477.2613L}) but
\citet{2019AJ....157...64L} showed that it is possible to compute them
analytically if the specific intensity is first expanded in a basis of
spherical harmonics and then mapped to a different basis more suitable for
evaluating the integrals.

@@

\section{Statistical inference -- theory}
\label{sec:inference_theory}
Having covered the basic physics behind gravitational lensing and occultation mapping I
now turn to the subject of statistical inference in general and more specifically --
Bayesian statistical inference. Bayesian (as opposed to frequentist)
provides a straightforward and elegant theoretical framework for solving inverse
problems of the kind we mentioned
in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}. These are problems
where we have little data per object of interest, the data is very noisy,
the physical models are relatively well understood but their parameters don't map
to observables in a straightforward way and there are multiple competing, sometimes
physically  quite different, models which provide a good explanation for the data.
Having said that, I am not particularly dogmatic about the use of Bayesian methods
and I will occasionally steer away from the classical approach and point out where
a less Bayesian approach is often superior in practice.  What is most important is the
use of \emph{probabilities} for encoding uncertainty about the physical world.

I start by reviewing basic probability theory and briefly mention the
frequentist interpretation. I then introduce two key concepts, the likelihood
function -- a recipe for generating plausible datasets given model parameters,
and priors -- what we know about the model parameters prior to gathering any
data. Next, I briefly summarize fundamental concepts such as the curse of
dimensionality and the no-free-lunch-theorem. Finally I end the section by
comparing machine learning to statistical inference within the physical
sciences.

\subsection{Probability theory}
There are two different interpretations of probability. In the
\emph{frequentist} interpretation probabilities are \emph{frequencies of
    events} which are repeated in a large number of trials (think of a repeated
coin tosses). An alternative to the frequentist interpretation is the
\emph{Bayesian} interpretation in which probabilities represent \emph{degrees
    of belief} about the subject of interest. In both cases probabilities are
necessarily positive numbers between zero and one. The major advantage of the
Bayesian interpretation is that it allows one to reason about the probability
of one-off events such as ``person X wins the election'' or ``this microlensing
event is a binary lens event''. The differences between the two interpretations
are not just philosophical because which interpretation one subscribes to can
determine how one approaches a particular statistical inference problem and
which sets of tools one uses. Both approaches are commonly used within the
physical sciences depending on the field and the problem at hand. For instance,
in case of particle physics frequentist methods are much more common because
particle physics experiments often involve billions of repeated experiments in
particle accelerators. One the other hand, Bayesian methods are more popular in
astronomy because we cannot do true repeated experiments with astrophysical
objects.

The mathematical rules for manipulating probabilities are quite
straightforward. In probability theory we use \emph{random variables} which can
be either discrete or continuous and their possible values are described with
probability distributions over all possible values the variable is allowed to
take. For discrete random variables we can talk of probabilities that a given
variable assumes a particular value, say $5$, but with continuous random
variables we can only talk of probabilities that a given variable is in a
particular range of values, say $[0,1]$. Now let's consider a continuous
\emph{random variable} $x$ with an associated probability density function
(pdf) $p(x)$\footnote{The notation $p(x)$ we use is somewhat overloaded. In
    statistics, the proper notation for a probability density function is
    $p_{X}(x)$ where $X$ is a random variable and $x$ is a particular realization
    of that random variable.} Since $x$ has to take on \emph{some values} within
its domain we have the requirement that
\begin{equation}
    \int p(x)\,\ud x=1
    \label{eq:normalization_condition}
\end{equation}
where the integral is taken over the entire domain of $x$.
In order for Equation~\ref{eq:normalization_condition} to hold,
$p(x)$ has to have units of $x^{-1}$ so that
the integral is a dimensionless number.

In general, we are usually dealing with several parameters at once in which
case we are interested in \emph{joint} probability distributions over all the
parameters. Consider random variables $x_1$ and $x_2$ with a joint probability
density function $p(x_1,x_2)$ which when integrated over a particular interval
gives the probability that both $x_1$ and $x_2$ are contained within that
interval. These joint probabilities can be decomposed into \emph{conditional
    probabilities} using the so called product rule:
\begin{align}
    p(x_1,x_2) & =p(x_1)\,p(x_2\lvert x_1) \\
    p(x_1,x_2) & =p(x_1\lvert x_2)\,p(x_2)
    \label{eq:product_rule}
\end{align}
where $p(x_2\lvert x_1)$ is a pdf for $x_2$ \emph{conditional on
    $x_1$ assuming a certain value.} $p(x_2\lvert x_1)$ is every bit as
valid pdf as $p(x_2)$, it has the same units and it has to integrate to 1.
Combining \ref{eq:product_rule} results in the famous \emph{Bayes's theorem}:
\begin{equation}
    p(x_1\lvert x_2)= \frac{p(x_2\lvert x_1)\,p(x_1)}{p(x_2)}
    \label{eq:bayes_theorem}
\end{equation}
Bayes's theorem is simply a rule for converting one kind of a conditional probability
into another.
Given a pdf $p(x_1,x_2)$, we can \emph{integrate out} or
\emph{marginalize} parameters that we are not interested in to obtain a distribution over
parameters of interest\footnote{
    It is very common in astronomy to have a model
    which consists of a set of parameters $\boldsymbol x_1$ of physical signal of interest and another set
    of ``nuisance parameters'' $\boldsymbol x_2$ which often describe the noise properties. In those cases
    we need to marginalize over parameters $\boldsymbol x_2$ to obtain a distribution over $x_1$.}:
\begin{equation}
    p(x_1)=\int p(x_1,x_2)\textrm{d}x_2=\int p(x_1\lvert
    x_2)\,p(x_2)\textrm{d}x_2
\end{equation}
where we applied Bayes's theorem in the second inequality.
In cases where the parameter space is high dimensional these integrals might be
impossible to calculate analytically but they can (sometimes) be estimated using sampling
algorithms as we shall se in subsequent sections.
In some cases, we may be able to factor out $p(x_1,x_2)$ as
If $x_1$ and $x_2$ are \emph{statistically independent} we can factor out $p(x_1,x_2)$ as
\begin{equation}
    p(x_1,x_2)=p(x_1)\,p(x_2)\,.
\end{equation}
Independence is a very common assumption in statistical modelling. For instance, when modeling light curves
of microlensing events or planetary transits we often assume that flux $f_i$ measured at time $t_i$ is independent
of flux $f_j$ measured at time $t_j$  for all measurement times $t_i,t_j$ so that
\begin{equation}
    p(\{f\}_{i=1}^N)=\prod_{i=1}^Np(f_i)
    \label{eq:likelihood_indep}
\end{equation}
If the pdf-s $p(f_i)$ also happen to be equal, we say that the data is independent and
identically distributed (``iid''). If the data is independent but not identically distributed
we say that the uncertainties are \emph{heteroscedastic}.
The iid assumption is not that common in astronomy because we tend to have some idea of per-data-point
uncertainties.

\subsubsection{Expectation values and the mode}
The \emph{mean} or the \emph{expected value} of a distribution $p(\vec{x})$,
usually denoted by $\mu$, is a measure of the ``center'' of a distribution and
it is defined as
\begin{equation}
    \mathbb{E}[x] \equiv\int x \,p(x) \ud x
\end{equation}
Similarly, the \emph{variance} is a measure of the ``spread'' of a distribution:
\begin{align}
    \mathbb{V}[x] & \equiv\mathbb{E}\left[(x-\mu)^{2}\right]=\int(x-\mu)^{2} p(x) d x                                           \\
                  & =\int x^{2} p(x) \ud x+\mu^{2} \int p(x) \ud x-2 \mu \int x p(x) \ud x=\mathbb{E}\left[x^{2}\right]-\mu^{2}
\end{align}
The variance is usually denoted by $\sigma^2$ and the square root of $\sigma^2$ is the
\emph{standard deviation}.
For a \emph{multivariate} random variable with $D$ dimensions then the variance generalizes to
a symmetric, positive semi definite matrix called the \emph{covariance matrix}:
\begin{align}
    \operatorname{Cov}[\vec{x}] & \equiv\mathbb{E}\left[(\vec{x}-\mathbb{E}[\vec{x}])(\vec{x}-\mathbb{E}[\vec{x}])^{\top}\right] \equiv \boldsymbol{\Sigma}                                                                                                                                                                                \\
                                & =\left(\begin{array}{cccc}
                                             \mathbb{V}\left[x_{1}\right]                & \operatorname{Cov}\left[x_{1}, x_{2}\right] & \cdots & \operatorname{Cov}\left[x_{1}, x_{D}\right] \\
                                             \operatorname{Cov}\left[x_{2}, x_{1}\right] & \mathbb{V}\left[x_{2}\right]                & \cdots & \operatorname{Cov}\left[x_{2}, x_{D}\right] \\
                                             \vdots                                      & \vdots                                      & \ddots & \vdots                                      \\
                                             \operatorname{Cov}\left[x_{D}, x_{1}\right] & \operatorname{Cov}\left[x_{D}, x_{2}\right] & \cdots & \mathbb{V}\left[x_{D}\right]
                                         \end{array}\right)
\end{align}

In the general case, we can compute expectation values of an arbitrary function
$g(\vec{x})$ under the probability distribution $p(\vec{x})$:
\begin{equation}
    \mathbb{E}\left[g(\vec{x})\right]=\int g(\vec{x})\,p(\vec{x})\ud x
    \ud\theta
\end{equation}
Some examples of $g$ include higher moments of the distribution, the median and
quantile estimates. The \emph{mode} of $p(\vec{x})$
\begin{equation}
    \vec{x}^{*}=\underset{\vec{x}}{\operatorname{argmax}}\,p(\vec{x})
\end{equation}
on the other hand is not an expectation value. It is the global maximum of the
function $p(\vec{x})$.
Estimating the mode involves computing the \emph{derivatives} of the density
$p(\vec{x})$, whereas expectation values require computing
\emph{integrals} over the space $\vec{x}$. Computing either
the mode or expectation values
for nontrivial densities $p(\vec{x})$ is a very hard and often impossible problem
(more on this in Section~\ref{sec:inference_practice}).

\subsubsection{Central Limit Theorem}
Consider $N$ \emph{independent and identically distributed} random variables
with pdf's $p_n(x)$ which are not necessarily Gaussian. Let $S_N$ be the sum
the random variables. The \emph{Central Limit Theorem} (CLT) says that as $N$
increases, the distributions of this sum approaches the Gaussian distribution
\begin{equation}
    p\left(S_{N}=u\right)=\frac{1}{\sqrt{2 \pi N \sigma^{2}}} \exp \left(-\frac{(u-N \mu)^{2}}{2 N \sigma^{2}}\right)
\end{equation}
and the distribution of the quantity
\begin{equation}
    Z_{N} \equiv \frac{S_{N}-N \mu}{\sigma \sqrt{N}}=\frac{\bar{x}-\mu}{\sigma / \sqrt{N}}
\end{equation}
where $\bar{x}$ is the \emph{sample mean} $\frac{1}{N}\sum_{i=1}^Nx_i=S_N/N$
converges to the Gaussian distribution with zero mean and unit variance.
Because lots of random variables representing real world processes can be seen as sums
of i.i.d. independent random variables the Gaussian distribution is all over the place
in statistics. One should keep in mind that the assumptions underlying the CLT are
not always satisfied and also the convergence towards a Gaussian is
more rapid in the bulk of the distribution around the mean than in the tails of the
distribution.

\subsubsection{Change of variables}
Given a univariate random variable $x$ the effect of some deterministic
transformation $y(x)$ is to stretch and shift the distribution $p(x)$. For an
infinitesimal interval the \emph{probability mass} is the same in both
variables so $p(x) \ud x=p(y) \ud y$ which yields the \emph{change of variables
    formula}
\begin{equation}
    p_{y}(y)=p_{x}(x)\left|\frac{\ud x}{\ud y}\right|
\end{equation}
In the multivariate case, if $\vec{f}$ is an invertible function mapping $\mathbb{R}^n$
to $\mathbb{R}^n$ with inverse $\vec{g}$, the pdf of $\vec{y}=\vec{f}(\vec{x})$ is
\citep{murphy_book_2022}
\begin{equation}
    p_{y}(\vec{y})=p_{x}(\vec{g}(\vec{y}))\left|\operatorname{det}\mathbf{J}_{g}(\vec{y})\right|
\end{equation}
where $\vec{J}_{g}=\frac{\ud \vec{g}(\vec{y})}{\ud \vec{y}^{\top}}$ is the Jacobian of $\vec{g}$
and $\operatorname{det}|\mathbf{J}(\vec{y})|$ is the determinant of the Jacobian.

\subsubsection{Kullback Leibler divergence}
A common way of measuring ``similarity'' between two distributions $p(x)$ and
$q(x)$ is using the \emph{Kullback-Leibler divergence} (KL divergence for
short):
\begin{equation}
    D_{\mathrm{KL}}(p \| q)=\int_{-\infty}^{\infty} p(x) \ln \left(\frac{p(x)}{q(x)}\right) \ud x
    \label{eq:kl_divergence}
\end{equation}
where $D_{\mathrm{KL}}(p \| q)$ should be read as ``the KL divergence from p to q''.
This is not a measure of ``distance'' between the two distributions because
it is not a symmetric quantity. The KL divergence is best understood as a quantity which
quantifies \emph{how surprised one would be if one expected to encounter distribution $p(x)$
    but instead saw distribution
    $q(x)$}\footnote{\url{https://wiki.santafe.edu/images/a/a8/IT-for-Intelligent-People-DeDeo.pdf}}.
This quantity is important in the context of variational inference which is covered in
Section~\ref{sec:inference_practice}.

\subsection{Bayes' theorem in practice}
\label{ssec:likelihood_function}
Consider a general inference problem where we collect some data $\mathcal{D}$ and we are interested
in the joint probability distribution for parameters $\boldsymbol\theta$ of particular model.
We can then rewrite Bayes' theorem as
\begin{equation}
    p(\vec{\theta} | \mathcal{D})=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{p(\mathcal{D})}=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{\int p\left(\vec{\theta}^{\prime}\right) p\left(\mathcal{D} |\vec{\theta}^{\prime}\right) d \vec{\theta}^{\prime}}
    \label{eq:bayes_theorem2}
\end{equation}
In Equation~\ref{eq:bayes_theorem2} the term $p(\mathcal{D}|\vec{\theta})$  is a probability
distribution over the data $\mathcal{D}$ given the parameters of the model $\vec{\theta}$.
When this term is seen as a function of the parameters given a fixed dataset, it is
called the \emph{likelihood function }  or the \emph{likelihood} for short. $p(\boldsymbol\theta)$
is to so called \emph{prior} distribution over $\vec{\theta}$ and it represents our beliefs
about $\vec{\theta}$ prior to having observed the data $\mathcal{D}$. When the prior distribution is
multiplied by the likelihood we obtain, up to a normalizing constant $p(\mathcal{D})$, the posterior
distribution over $\vec{\theta}$ given the data $\mathcal{D}$.
The best way to think about Equation~\ref{eq:bayes_theorem2} is seeing at as a rule
for updating beliefs. Our prior beliefs about the parameters of the model
$p(\mathrm{parameters})$ should be updated to posterior beliefs $
    p(\mathrm{parameters} | \mathrm{data})$ by making use of the likelihood
$p(\mathrm{data} | \mathrm{parameters})$. If the data is useful, the likelihood is generally
much narrower than the prior pdf and the choice of priors does not significantly
influence the final results.

\subsubsection{The likelihood}
Let's cover each of these three terms in more detail, starting with the most
important one -- the likelihood. The likelihood can be seen as a \emph{recipe
    for generating plausible datasets $\mathcal{D}$} (a \emph{generative model})
given a specific realization of the parameters $\vec{\theta}$. For the kinds of
models we work with in astronomy, the likelihood consists of two parts. The
first part is (usually deterministic) function which describes the physical
system of interest. This is for instance the predicted flux for a binary
microlensing event with a particular trajectory. The second part is the
\emph{noise model} which describes the statistical properties of the noise
inherent in any physical measurement. We often assume that the noise is
independent and Gaussian. It is very common to use the likelihood as is and
optimize it with respect to the model parameters $\vec{\theta}$ in a procedure
called \emph{maximum likelihood estimation} (MLE) to find a single point in the
parameter space $\vec{\theta}^\star$ which maximizes the likelihood (more on
this in Section~\ref{ssec:mle_estimation}).
